Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
paper full

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$.\MANUEL{necesitamos las ecuaciones? Swap de una permutación es simplemente la distancia de kendall de la permutación inversa.} %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

@@

%During our experiments, we realized that the \emph{swap} distance used by CEGO
%assumes rankings, i.e., it actually measures the \ken distance, yet CEGO did
% not invert the permutation before evaluating the PFSP.
During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP.\MANUEL{no me convence esta forma de expresarlo porque la crítica obvia sería que por qué invertimos la función objectivo y no la función de distancia?} We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. Of course, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. Therefore, in the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

@@

In the following, we compare 5 algorithm configurations. Three configurations
of UMM (UMM$_{100}$, UMM$_{200}$, and UMM$_{400}$) that differ in their budget
of evaluations ($m$). As mentioned in Sec.~\ref{sec:sampling}, the behavior of
UMM depends on the value of $m$. The other two configurations correspond 
to \CEGOorig and \CEGOinv, as defined above.

Since there is a small number of evaluations and the algorithms evaluate just
one solution at each step, we chose to record and plot the fitness of the
solution evaluated at each step of each run, instead of the usual plots of the
best-so-far solution over number of evaluations. We believe that plotting each
evaluated solution provides better insights on the search dynamics of the
algorithms analyzed here. Please note that the algorithms do not return the
final solution shown in the plots, but rather the best one (minimum fitness) of
each run. Nevertheless, the minimum values achieved in the plot up to a
particular number of function evaluations gives an estimation of the mean
fitness of the best-so-far solution up to that point in the run. We provide
later an analysis of the best solution found by each run of the algorithm.

@@

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
configurations of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance of the model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets. \manuel{The
average runtime for all the instances is around 0.3 minutes, 1.15 minutes and
4.5 minutes respectively for the versions with 100, 200 and 400 fitness
evaluations.}{}\MANUEL{We should discuss the time when discussing the tables.}
\EKHINE{luego el detalle, que la gente no llega al final }

@@

\myparagraph{Comparison of the best solutions found} We now turn our analysis
to the best (minimum) fitness found at the end of the run by each algorithm. We
show in Table~\ref{tab:results} the mean fitness of the best solution found
over the 10 runs of each algorithm.

For the reason for the abrupt change between both version for
every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the
original version, there is a global decrease in the fitness function. However,
a closer analysis of each iteration does not show a real convergence to any
point. However, with the modified version in which rankings are modeled and
orderings are evaluated we can appreciate the a decrease in terms of fitness
per evaluation. Overall, the modified CEGO has a very abrupt decrease at the
first 50 iterations. At one point, the algorithm seems to have converged and
very little improvement is don from the 150-th iteration onwards.  The average
running time is approximately 22 hours for the 400 function evaluations, which
is 300 times slower than UMM for the same number of function evaluations.

\manuel{%
This result summarization is not the classical. However, due to the limited number of function evaluations, we can have a better picture of the behaviour of the algorithm at each step. As a consequence, results displayed in the figures are not comparable with the original paper of CEGO: for this comparison we provide in Table~\ref{table:fitness} the summary of the best solutions found by each algorithm, which are similar to the original study. }{}

\paragraph{Figure descriptiopn} Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
\begin{itemize}
\item UMM\_100, UMM\_200 and UMM\_400 are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM\_100, UMM\_200 wont reach the right corner of the figure. 
\item CEGO\_orig and CEGO\_inv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
\end{itemize}

\paragraph{Analysis} We start our analysis with the three version of UMM. We can see a general trend for all the instances: After the 10 random initial permutations, the fitness of the new solutions decreases with a linear trend along iterations. The (close to) linear decrease is given by the linear decrease in the expected distance of the model which is sampled at each step, described in Section~\ref{sec:sampling}. Overall, the improvement does not seem to converge not even for the longer runs of 400 evaluations.
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations.

The red and the gold lines correspond to CEGO, the original and our modified versions. For the reason for the abrupt change between both version for every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the original version, there is a global decrease in the fitness function. However, a closer analysis of each iteration does not show a real convergence to any point. However, with the modified version in which rankings are modeled and orderings are evaluated we can appreciate the a decrease in terms of fitness per evaluation. Overall, the modified CEGO has a very abrupt decrease at the first 50 iterations. At one point, the algorithm seems to have converged and very little improvement is don from the 150-th iteration onwards.  The average running time is approximately 22 hours for the 400 function evaluations, which is 300 times slower than UMM for the same number of function evaluations.

The performance of the UMM is clearly better than the performance of the original CEGO for the same problem. For CEGO\_inv on the other hand improves the results of the UMM with the current experimental setting. However, UMM seems to be improving the quality of the solutions until the last iteration while CEGO\_inv seems to be stucked at a local optimum. Further analysis lines will consider a larger budget of fitness evaluation.

Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p50-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

\paragraph{Discussion on other problems: the case of QAP}

@@

The overall conclusion from Table~\ref{tab:results} is that there is not a
significant difference in performance between UMM and CEGO. For the larger LOP
instances, CEGO reaches the time limit of 144 hours before consuming the budget
of 400 evaluations and thus, obtains quite poor solutions compared to UMM,
which is at least an order of magnitude faster.

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found. % at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of UMM minus the one of CEGO. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
  \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-p50-02     & 19090.5                      & (285.4)                                                                         & 21746.6   & (645.4)   & {[}-3140.6,& -2171.6{]}    & 22.9 & 0.5 \\
         & N-p50-01     & 18197.7                      & (225.6)                                                                         & 21161.8   & (657.4)   & {[}-3447.4,& -2480.8{]}    & 23.2 & 0.5 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}}
\end{table*}

\begin{table*}[tb]%time
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &  CEGO\_inv &  CEGO\_orig &  UMM\_400 &  UMM\_200 &  UMM\_100 \\
\midrule
    LOP &  N-t59b11xx &   1265.98 &    1240.36 &    22.24 &     5.78 &     1.37 \\
    LOP &  N-t59d11xx &   1269.93 &    1252.89 &    21.84 &     5.62 &     1.37 \\
    LOP &    N-p40-01 &   1195.90 &    1170.81 &    18.13 &     4.61 &     1.14 \\
    LOP &    N-p40-02 &   1207.24 &    1183.09 &    18.11 &     4.76 &     1.15 \\
    LOP &    N-p50-01 &   1391.55 &        NaN &    27.36 &      NaN &      NaN \\
    LOP &    N-p50-02 &   1372.71 &        NaN &    28.36 &      NaN &      NaN \\
    LOP &  N-sgb75.01 &   1994.22 &    2004.70 &    62.75 &    15.77 &     3.97 \\
    LOP &  N-sgb75.02 &   1995.09 &    1975.41 &    64.58 &    15.90 &     4.03 \\
   PFSP &       rec05 &   1099.63 &     917.81 &     4.54 &     1.14 &     0.30 \\
   PFSP &       rec13 &   1109.51 &    1043.08 &     4.55 &     1.07 &     0.31 \\
   PFSP &       rec19 &   1090.06 &    1087.38 &    10.51 &     2.59 &     0.63 \\
   PFSP &       rec31 &   1387.09 &    1367.26 &    27.84 &     6.90 &     1.79 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[tb]%fitness
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &   CEGO\_inv &   CEGO\_orig &   UMM\_400 &   UMM\_200 &   UMM\_100 \\
\midrule
    LOP &  N-t59b11xx &          102408.0 &           126204.0 &         114075.5 &         115941.5 &         128195.5 \\
    LOP &  N-t59d11xx &           12692.2 &            30199.8 &          16662.9 &          22117.4 &          28050.4 \\
    LOP &    N-p40-01 &           10496.6 &            15027.7 &          12369.1 &          13043.7 &          14459.3 \\
    LOP &    N-p40-02 &           11714.7 &            15150.2 &          13160.5 &          13923.2 &          15027.7 \\
    LOP &    N-p50-01 &           18197.7 &                NaN &          21161.8 &              NaN &              NaN \\
    LOP &    N-p50-02 &           19090.5 &                NaN &          21746.6 &              NaN &              NaN \\
    LOP &  N-sgb75.01 &         1893533.2 &          2095899.6 &        1992059.9 &        2011271.7 &        2074111.6 \\
    LOP &  N-sgb75.02 &         1895801.8 &          2069286.9 &        1967132.3 &        2011004.2 &        2045556.5 \\
   PFSP &       rec05 &            1276.2 &             1318.7 &           1276.3 &           1283.1 &           1300.4 \\
   PFSP &       rec13 &            1992.9 &             2096.0 &           2057.8 &           2113.6 &           2134.9 \\
   PFSP &       rec19 &            2264.5 &             2328.4 &           2304.8 &           2346.5 &           2392.4 \\
   PFSP &       rec31 &            3415.5 &             3512.3 &           3419.3 &           3438.9 &           3511.9 \\
\bottomrule
\end{tabular}
\end{table*}

@@

\section{Conclusions}\label{sec:conclusions}

In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $100$, $200$ and $400$ evaluations).
Our experiments presented here show that UMM is able to match the results of
the state-of-the-art Bayesian optimizer for expensive black-box
optimization if the computation time of the optimizers is not a concern.

@@

In this paper, we discuss the importance of the representation of permutations in applied mathematics problems. In the context considered in this paper, permutations can bee seen as rankings or orderings of items. In other problems they could be matchings or cycles. 
While certain functions are
this discussion is usually ommited in the literature.
We discuss the different concepts and show, experimentally, the consequences of misunderstanding both.

>>>>

In Section~\ref{sec:permutations}, we highlighted the important difference between rankings and orderings, usually neglected in the optimization literature. Statistical models (Borda or the \ken distance) are usually defined for rankings while the optimization literature usually builds upon orderings. For example, the Swap distance $d_s(\sigma,\pi)$ is the \ken distance counterpart in optimization, $d_s(\sigma,\pi) = d(\sigma^{-1}, \pi^{-1})$.\MANUEL{necesitamos las ecuaciones? Swap de una permutación es simplemente la distancia de kendall de la permutación inversa. .}
\EKHINE{asi mas claro, no se van a perder con esto}
 %The concepts of rankings and orderings are usually confused of ignored. An example of such confusion is the claim that Swap distance is the same metric as  the \ken distance, which is false except for particular permutations. In reality, the \emph{swap} distance is the \ken distance of the inverted permutations.

@@

%During our experiments, we realized that the \emph{swap} distance used by CEGO
%assumes rankings, i.e., it actually measures the \ken distance, yet CEGO did
% not invert the permutation before evaluating the PFSP.
During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP.\MANUEL{no me convence esta forma de expresarlo porque la crítica obvia sería que por qué invertimos la función objectivo y no la función de distancia?} \EKHINE{es lo mismo}We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. Of course, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. Therefore, in the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

@@

%In the following, we compare 5 algorithm configurations.
% Three configurations
%of UMM (UMM$_{100}$, UMM$_{200}$, and UMM$_{400}$) that differ in their budget
%of evaluations ($m$). As mentioned in Sec.~\ref{sec:sampling}, the behavior of
%UMM depends on the value of $m$. The other two configurations correspond 
%to \CEGOorig and \CEGOinv, as defined above.

Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
\begin{itemize}
\item UMM\_100, UMM\_200 and UMM\_400 are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM\_100, UMM\_200 wont reach the right corner of the figure. As mentioned in Sec.~\ref{sec:sampling}, the behavior of
UMM depends on the value of $m$. 
\item CEGO\_orig and CEGO\_inv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
\end{itemize}

Since there is a small number of evaluations and the algorithms evaluate just
one solution at each step, we chose to record and plot the fitness of the
solution evaluated at each step of each run, instead of the usual plots of the
best-so-far solution over number of evaluations. We believe that plotting each
evaluated solution provides better insights on the search dynamics of the
algorithms analyzed here, despite it is not directly comparable with most literature results. Please note that the algorithms do not return the
final solution shown in the plots, but rather the best one (minimum fitness) of
each run. Nevertheless, the minimum values achieved in the plot up to a
particular number of function evaluations gives an estimation of the mean
fitness of the best-so-far solution up to that point in the run. We provide
later an analysis of the best solution found by each run of the algorithm, which are comparable with the original paper of CEGO.

@@

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
configurations of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance of the model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets.

@@

\myparagraph{Comparison of the best solutions found}
%
We now turn our analysis to the best (minimum) fitness found at the end of the
run by the two best algorithms, that is, \CEGOinv and UMM$_{400}$. We show in
Table~\ref{tab:results} the mean fitness (and standard deviation) of the best
solution found over the 10 runs of each algorithm, together with a 95\%
confidence interval (CI) around their mean fitness difference. Except for PFSP
instances \texttt{rec05} and \texttt{rec31}, the differences are statistically
significant in favour of \CEGOinv (the intervals do not contain the value
zero). Nevertheless, the CIs indicate that the differences, although
consistent, are relatively small.

\myparagraph{Runtime comparison}
%
The average
running time is approximately 22 hours for the 400 function evaluations, which
is 300 times slower than UMM for the same number of function evaluations.

Table~\ref{tab:results} shows a brief comparison of time and the full ones are given in Table~\ref{tab:time}. The difference is time in impressive. 
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
For CEGO in both versions the average time for 400 evaluations is more than 22 hour. This is 300 times slower than UMM for the same number of function evaluations. 
Moreover, UMM scales good for larger instances in quality performance. 
The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \ekhine{MANUEL, tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-04} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-sgb75-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

%
%\paragraph{Figure description} Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
%\begin{itemize}
%\item UMM\_100, UMM\_200 and UMM\_400 are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM\_100, UMM\_200 wont reach the right corner of the figure. 
%\item CEGO\_orig and CEGO\_inv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
%\end{itemize}
%
%\paragraph{Analysis} We start our analysis with the three version of UMM. We can see a general trend for all the instances: After the 10 random initial permutations, the fitness of the new solutions decreases with a linear trend along iterations. The (close to) linear decrease is given by the linear decrease in the expected distance of the model which is sampled at each step, described in Section~\ref{sec:sampling}. Overall, the improvement does not seem to converge not even for the longer runs of 400 evaluations.
%The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
%
%
%The red and the gold lines correspond to CEGO, the original and our modified versions. For the reason for the abrupt change between both version for every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the original version, there is a global decrease in the fitness function. However, a closer analysis of each iteration does not show a real convergence to any point. However, with the modified version in which rankings are modeled and orderings are evaluated we can appreciate the a decrease in terms of fitness per evaluation. Overall, the modified CEGO has a very abrupt decrease at the first 50 iterations. At one point, the algorithm seems to have converged and very little improvement is don from the 150-th iteration onwards.  The average running time is approximately 22 hours for the 400 function evaluations, which is 300 times slower than UMM for the same number of function evaluations. 
%
%The performance of the UMM is clearly better than the performance of the original CEGO for the same problem. For CEGO\_inv on the other hand improves the results of the UMM with the current experimental setting. However, UMM seems to be improving the quality of the solutions until the last iteration while CEGO\_inv seems to be stucked at a local optimum. Further analysis lines will consider a larger budget of fitness evaluation. 
%
%Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
%The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.
%

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of \CEGOinv minus the one of UMM$_{400}$. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
 % \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}
%}
\end{table*}

@@

\begin{table*}[tb]%time
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &  CEGO\_inv &  CEGO\_orig &  UMM\_400 &   UMM\_200 &  UMM\_100 \\
\midrule
    LOP &    N-p40-01 &   1195.90 &    1170.81 &     3.12 &  0.850000 &     0.21 \\
    LOP &    N-p40-02 &   1207.24 &    1183.09 &     3.11 &  0.830000 &     0.20 \\
    LOP &  N-t59b11xx &   1265.98 &    1240.36 &     3.78 &  0.950000 &     0.23 \\
    LOP &  N-t59d11xx &   1269.93 &    1252.89 &     3.93 &  0.966667 &     0.25 \\
    LOP &  N-sgb75.01 &   1994.22 &    2004.70 &    10.88 &  2.730000 &     0.72 \\
    LOP &  N-sgb75.02 &   1995.09 &    1975.41 &    10.88 &  2.760000 &     0.69 \\
   PFSP &       rec05 &   1099.63 &     917.81 &     0.80 &  0.200000 &     0.10 \\
   PFSP &       rec13 &   1109.51 &    1043.08 &     0.85 &  0.200000 &     0.10 \\
   PFSP &       rec19 &   1090.06 &    1087.38 &     1.76 &  0.470000 &     0.10 \\
   PFSP &       rec31 &   1387.09 &    1367.26 &     4.77 &  1.280000 &     0.32 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[tb]%fitness
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &   CEGO\_inv &  CEGO\_orig &    UMM\_400 &       UMM\_200 &    UMM\_100 \\
\midrule
    LOP &    N-p40-01 &    10496.6 &    15027.7 &    12453.7 &  1.336170e+04 &    14649.9 \\
    LOP &    N-p40-02 &    11714.7 &    15150.2 &    13191.9 &  1.420900e+04 &    15206.2 \\
    LOP &  N-t59b11xx &   102408.0 &   126204.0 &   114726.5 &  1.202020e+05 &   132039.5 \\
    LOP &  N-t59d11xx &    12692.2 &    30199.8 &    16775.4 &  2.352089e+04 &    29884.1 \\
    LOP &  N-sgb75.01 &  1893533.2 &  2095899.6 &  1983521.2 &  2.004006e+06 &  2046895.1 \\
    LOP &  N-sgb75.02 &  1895801.8 &  2069286.9 &  1973083.9 &  2.001244e+06 &  2039356.1 \\
   PFSP &       rec05 &     1276.2 &     1318.7 &     1282.1 &  1.287000e+03 &     1320.2 \\
   PFSP &       rec13 &     1992.9 &     2096.0 &     2077.7 &  2.108500e+03 &     2131.6 \\
   PFSP &       rec19 &     2264.5 &     2328.4 &     2318.5 &  2.352100e+03 &     2386.5 \\
   PFSP &       rec31 &     3415.5 &     3512.3 &     3420.5 &  3.490900e+03 &     3474.3 \\
\bottomrule
\end{tabular}\end{table*}

@@

\section{Conclusions}\label{sec:conclusions}
In this paper, we have introduced UMM, a population-based probabilistic
algorithm based on an unbalanced Mallows model. The algorithm is designed for
black-box combinatorial problems on permutation landscapes and when the budget
of fitness evaluations is severely limited (here, up to $100$, $200$ and $400$ evaluations).
Our experiments presented here show that UMM is able to match the results of
the state-of-the-art Bayesian optimizer for expensive black-box
optimization if the computation time of the optimizers is not a concern.

@@

In this paper, we discuss the importance of the representation of permutations in applied mathematics problems. In the context considered in this paper, permutations can bee seen as rankings or orderings of items. In other problems they could be matchings or cycles. 
While certain functions are
this discussion is usually omitted in the literature.
We discuss the different concepts and show, experimentally, the consequences of misunderstanding both.

Although we were inspired by previous work on ACO for expensive black-box
combinatorial problems~\citep{PerLopStu2015si}, UMM is, to the best of our
knowledge, the first EDA specifically designed for such problems that is able
to match (and sometimes even improve) the state-of-the-art
optimizer. Moreover, it is overwhelmingly faster than the alternatives for
large problem sizes.  The results presented here show that, despite the
intrinsic difficulty of the expensive black-box combinatorial scenario, there
are still significant advances yet to be made. Thus, we hope that this paper
will motivate further research.

In this paper, we focus on the UMM under the \ken distance. Therefore, we consider appropriate problems for it, such as the Linear Ordering Problem and the Permutation Flow-shop Scheduling Problem, where the pairwise ordering among the items is relevant (as opposed to the actual exact position of each item in the permutation)
